Improving the reliability of cognitive task measures: A narrative review

### 2022-10-26 manuscript/main.tex
checkpointing progress

Cognitive tasks have already demonstrated, and continue to hold, tremendous promise for biological psychiatry. Here we define cognitive task as any experimental paradigm that measures some behavior in order to make inferences about one or more cognitive processes (e.g. Stroop task, delay discounting task, reversal learning task). Variability in performance on cognitive tasks provide experimenters with crucial insights into how individual differences in cognitive processing and ability relate to behavioral phenomena outside of the laboratory or clinic (e.g. reward sensitivity in a monetary-incentivized task and anhedonia in depression). Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{peng2019meta, spiegel2021relations} and cognitive decline \cite{salthouse2010selective}. Cognitive tasks then are invaluable tools for building and refining our understanding of psychiatric symptoms and syndromes. For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

% P2: Three key properties of behavioral measures: discriminability, reliability, and validity

The psychometric quality of a cognitive task measure (i.e. any descriptive or model-based indicator of behavioral performance such as proportion correct, average response time, and learning rate) for characterizing individual differences are summarized by three key properties: discriminating power, reliability, and validity \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to produce variability in participants' performance. This is a necessary property of tasks to be used for studying individual differences; where there is no variation in performance, there are no individual differences to study. The reliability of a task characterizes the degree to which it consistently rank-order participants across measurements. Put another way, a task measure is reliable if, assuming participants have not changed, it produces the same score for each participant over time. Finally, the validity of a task measure concerns whether it actually measures what it purports to measure. The focus of the current review is task reliability, which is a prerequisite for validity.

% P3: definition and Importance of reliability

What is the formal definition of reliability? In classical test theory, the variance in observed scores on a task measure (X) is the sum of the true score variance (T), reflecting real individual differences in the latent construct of interest, and measurement error (E; i.e. X = T + E). The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance (i.e. R = T/[T + E]). In other words, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences and the less it reflects noise.. It should now be clearer why reliability is a prerequisite for validity; when a task measure is unreliable, it reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g. task measure and self-reported symptom score) is a function of their individual reliabilities \cite{Spearman1904-mo}:

\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}

where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As can be observed in the equation above, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure. As reliability decreases, so too does the maximum observable correlation between itself and any other measure (Figure 1). As an important corollary, as reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

@@

To further complicate matters, the reliability of a task measure is not universal. The reliability of a task reflects interactions between its design, the participants completing it, and the context in which it is administered. Indeed task reliability can vary as a function of experimental design (stimulus sets, number of trials, time limits \cite{paap2016role, cooper2017role}); sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in-clinic, online); and response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); and scoring or estimation method \cite{Rouder2019-am, haines2020learning}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children; their task performance may drop to chance-level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may in turn lead to an increase in measurement error and a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. Ideally then, experimenters should investigate and report the reliability of task measures as part of any analysis or, at the very least, after having made changes to a task or scoring procedure or when administering the task to new sample populations or in new testing contexts.

@@

Though verifying the reliability of cognitive task measures is paramount to successful individual differences research, the reliability of task measures are historically seldom reported \cite{Green2016-xw, Parsons2019-jw}. When reported, it has frequently been observed that task measures exhibit lower reliability than what is conventionally considered acceptable for individual differences research (e.g. $\rho$ = 0.7 - 0.8) and lower than what is regularly achieved by self-report measures. Indeed many studies have now found that task measures to exhibit moderate-to-low reliability \cite{Hedge2018-lf, Frey2017-uz, Enkavi2019-oh, Von_Bastian2020-tm, Nitsch2022-pe, verdejo2021unified}. One possible explanation for this finding is the so-called ``reliability paradox'' of cognitive tasks \cite{hedge2018reliability}, which states that the often lackluster reliability of tasks is a result of a mismatch in goals between experimental and individual differences psychological research. In experimental psychology, the goal is often to demonstrate the existence of a behavioral effect. One sure-fire means of increasing the power to detect an effect is to minimize between-participants variance of a task measure. This of course is the exact opposite of what is desirable for individual differences research, where between-participants variance is essential to achieving reliable task measures. Thus, the tendency in biological psychiatry to adopt the most prominent tasks in experimental psychology --- the ones that most reliably demonstrate a behavioral effect --- may actually hamstring our efforts to study individual differences.

@@

Regardless we do not believe that task measures are inherently less reliable and that pessimism about task-based individual differences research is warranted. It is possible to (re)design tasks to achieve good reliability, even to the high levels dictated by conventional standards \cite{waltmann2022sufficient, sullivan2022enhancing, kucina2022solution, snijder2022psychometric}. The purpose of the current article is to provide a narrative review of approaches to improve task measure reliability. The article is divided into two main parts. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of approaches for improving task reliability using concrete examples from the published literature. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing task, to achieve sufficient reliability for use in individual differences research.

@@

We begin with a brief discussion of calculating reliability for cognitive task measures. There are two measures of reliability of relevant for task measures: internal consistency and test-retest reliability. Internal consistency is the reliability of a measure in a single administration. In turn, test-retest reliability (also known as temporal stability) is reliability of a task measure across two or more administrations. The two types of reliability provide complimentary information: internal consistency quantifies the consistency of individual difference on a task measure in a given testing session, whereas test-retest reliability quantifies the stability of individual differences across time. Even for experimenters primarily interested in cross-sectional experiments, test-retest reliability is a useful index. This is because estimates of internal consistency tend to be inflated due to between-participants variance from construct-irrelevant state-dependent factors (e.g. current mood, fatigue). Given sufficient time between testing sessions, test-retest reliability should be less biased state factors. A second point to consider with respect to these two indices is that, whereas high internal consistency is virtually always desirable for individual difference studies, high test-retest reliability may not be. Test-retest reliability should be interpreted with respect to the construct of interest. For example, low 1-month test-retest reliability may not be problematic for an index of a transient cognitive process (e.g. mood effects on cognition), but may be problematic if the effect is hypothesized to be stable across time (e.g. personality effects on cognition).

The test-retest reliability of a task measure can be calculated in numerous ways \cite{Parsons2019-jw}. Perhaps the simplest approach is to compute the Pearson correlation between participants' scores from two sessions. An alternative approach is to calculate the intraclass correlation coefficient (ICC), which decomposes a task measure into true score variance and error variance. There are many formulas for calculating ICC \cite{Koo2016-fz}, with the critical distinction being reliability is based on the \textit{consistency} or \textit{absolutely agreement} of a task measure across two administration. Consistency-based ICCs are affected only by the relative ordering of participants across time; that is, they are insensitive to systematic changes to the actual values of a task measure across time (e.g. practice effects). In contrast, absolute agreement-based ICCs measure the degree to which scores are identical across time. They are sensitive to changes in the overall mean of a task measure across sessions. The type of ICC to use depends on the experimenter's goals. If an experimenter cares only about the rank-ordering of participants across time, a consistency-based ICC (or Pearson correlation) is preferable. Conversely, if systematic differences in a task measure across task administrations is a concern, then an absolute agreement-based ICC should be preferred.

Calculating the internal consistency is more complicated for task measures. The most common measure of internal consistency, Cronbach's $\alpha$, requires certain assumptions that are unrealistic for task data (e.g. equivalence of trials, uncorrelated measurement error). As such, internal consistency for task measures is instead usually calculated via split-$K$ reliability instead, where $K$ indicates the number of splits. Split-half reliability ($K=2$) is most frequently chosen. A critical challenge in calculating split-half reliability is in deciding how to partition the data, as estimates of reliability may be biased when partitions are not equally good measures of a behavior \cite{Green2016-xw, pronk2022can}. For example, first-second splitting (i.e. partitioning the data by the first and second halves of an experiment) may be confounded by practice, fatigue, or other linear time effects. Under such conditions, first-second partitioning will spuriously decrease the similarity of data across partitions, thereby increasing estimates of measurement noise and underestimating reliability. In contrast, odd-even splitting (i.e. partitioning the data by odd and even trials) may be confounded when behavior across trials is non-independent (i.e. correlated errors). Under such conditions, odd-even splitting will artificially inflate the similarity of data across partitions, thereby decreasing estimates of measurement noise and overestimating reliability. Therefore, where possible, a permutation-based approach to calculating split-half reliability has been recommended \cite{pronk2022can, Parsons2019-jw}. Briefly, the idea is to compute the average reliability across many thousands of random partitions of the data into halves. For task measures derived from cognitive models, it may be unreasonably computationally intensive to employ such an approach. In which case, it may instead be preferable to design tasks with at least two independent blocks where the model is fit to each block independently and reliability is calculated using the model parameters estimated from each.

A final point is that, when calculating reliability for task measures, traditional sum or mean score estimates of performance (e.g. proportion correct responses, mean response time) may substantially underestimate reliability \cite{Rouder2019-am, haines2020learning}. These summary scores of task performance neglect trial-by-trial variability in performance, which can result in noisier estimates of task performance and consequently diminished estimates of reliability. Instead, it is often preferable to use trial-level hierarchical (mixed effects) models that can simultaneously estimate between- and within-participant (i.e. trial-level) variability in behavior. By directly modeling trial-by-trial variability, estimates of participants' performance are effectively denoised. Furthermore, the use of hierarchical models allows for partial pooling of trial data across participants, which further increases the precision of estimates of participants' performance. Together these properties can substantially boost estimates of reliability \cite{Rouder2019-am, haines2020learning}. In practice, multiple studies have demonstrated how the use of hierarchical trial-level models directly improves estimates of task measure reliability \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}. For a detailed discussion of hierarchical models in the context of task reliability, see Haines (this issue).

@@

As defined above, the reliability of a task measure is proportion of variance attributable to between-participant differences relative to measurement error. Thus, there are two major strategies for improving the reliability of a measure: by increasing between-participant variability or by decreasing measurement error. In what follows, we discuss approaches for accomplishing both strategies in turn. Where appropriate, we highlight studies from the published literature that are exemplary for imporiving the reliability of a task measure by implementing a particular strategy.

@@

By definition, the reliability of a task measure is zero when there is no variability across participants. Thus, range restriction of task measures via ceiling or floor effects is a serious obstacle to reliability. Siegelman and colleagues \cite{siegelman2017measuring} noted the consequences of floor effects on reliability in the context of statistical learning tasks. In statistical learning tasks, participants must learn to identify subtle patterns in the transition probabilities underlying a continuous sequence of stimuli. In reanalyzing archival datasets, Siegelman and colleagues found a majority of participants were at chance-level performance in discriminating between legitimate and foil sequence patterns; consequently, the reliability of conventional proportion correct measures suffered. In response, the authors designed a new version of a statistical learning task involving stimulus sequences that ranged more widely in their difficulty to learn. Only a minority of participants were at chance-level performance on this new task and, as such, the reliability of proportion correct scores improved. Similarly, in developing an abbreviated working memory task, Oswald and colleagues \cite{oswald2015development} found that they could remove the easiest trials --- those with ceiling level performance --- with virtually no change to task reliability. This is because working memory trials with performance at ceiling are incapable of differentiating ability across participants and therefore cannot contribute significantly to the reliability of the task.

Experimenters who are administering a task to a new population should be wary of range restriction effects. Cognitive tasks calibrated for one group of participants may be too easy or difficult for a different group. For example, Arnon and colleagues \cite{arnon2020current} found that statistical learning tasks developed for adults were too difficult for young children and therefore yielded unreliable discrimination scores in that population. Similarly, Kyllonen and colleagues \cite{kyllonen2019general} were motivated to develop a battery of fluid reasoning measures for highly educated adults due to observed ceiling effects in performance when using preexisting fluid reasoning tasks in this population. In sum, experimenters should be cautious when administering preexisting tasks to new populations; the tasks calibrated for one group may not be adequately sensitive for others.

@@

A related issue for task reliability is practice effects, or when participants' performance on a task improves with repeated administrations. Practice effects are relatively common for cognitive tasks \cite{hausknecht2007retesting, scharfen2018retest}. They might occur due to the attenuation of task-irrelevant nuisance factors (e.g. performance anxiety) and/or the learning of task-specific knowledge or strategies. Practice effects are not inherently an issue for reliability --- especially if an experimenter is only interested in the consistency, but not the absolute agreement, of participants' performance over time --- but they can become a pernicious issue if they are severe enough to induce ceiling effects. For example, Paredes and colleagues \cite{paredes2021psychometric} observed large practice effects on the Pavlovian go/no-go task for short retest intervals (3 days, 14 days), which resulted in poor estimates of test-retest reliability. In developmental and lifespan studies, practice effects are potentially complicated by their interaction with age \cite{anokhin2022age, salthouse2010influence}; that is, practice-induced ceiling effects may present in some age groups but not others.

One strategy for minimizing practice effects is simply to increase the time interval between task administrations. The more time that elapses between sessions, the greater the probability that participants have forgotten task-specific knowledge or strategies \cite{hausknecht2007retesting, scharfen2018retest}. Of course, this solution may not always be possible or desirable, especially if an experimenter is only able to or specifically interested in studying a behavior over a short time period.

A second strategy is to design tasks so as to prevent or discourage the formation of task-specific strategies. For example, McLean and colleagues \cite{mclean2018towards} investigated the repeatability of the beads task. In the beads task, participants are presented with two jars containing beads of two colors in equal but opposite ratios. In one trial of the task, a predetermined sequence of beads is drawn from one jar. After participants must decide which jar beads were being drawn from or request to see more beads. In a typical version of the task, the same sequence of bead draws is used across all trials. McLean and colleagues found that participants were aware that the sequence repeated across trials and, as a consequence, became more erratic in their decision to witness more bead draws with additional trials. In response, the authors developed a new version of the beads task that included distractor sequences of bead draws. This new design prevent participants from becoming aware of the target sequence, which in turn motivated participants to respond more consistently thereby improving the reliability of participants' information seeking scores.

@@

The preceding sections described potential threats to between-participants variability, but did not address how to actually increase it. A primary strategy for increasing between-participants variability (and therefore reliability) is to enhance the experimental effect. Amplifying the magnitude of an experimental effect (e.g. making a task more challenging, increasing the potency of a stress or affective induction) typically increases the range of participants' responses to it. Two recent studies highlight this approach in the context of cognitive control tasks.

Kucina and colleagues \cite{kucina2022solution} investigated the reliability of cognitive conflict effects in new versions of several standard cognitive control tasks (e.g. Stroop, Flanker, Simon) that amplified cognitive interference via two task design features. First, they combined multiple sources of cognitive interference in the same task (e.g. combining the Stroop and Simon effects to create a ``Stroopon'' task). Second, for a subset of trials, they required participants to make multiple responses based on both relevant and irrelevant stimuli attributes. Compared to the standard version of these tasks, these experimental manipulations had the effect of increasing task demands, which resulted in greater between-participants variance and reliability as a consequence. Similarly, Snijder and colleagues \cite{snijder2022psychometric} studied the reliability of cognitive conflict effects in multiple redesigned cognitive control tasks (e.g. Stroop, AX-CPT, Cued Task Switching, Sternberg) that required of participants either more proactive control (i.e. maintained vigilance) or reactive control. The authors found that the proactive control versions of the task were more demanding, which similarly led to larger cognitive interference effects and ultimately task reliability.

For decision-making tasks, or other tasks requiring participants to choose between two or more discrete options, a related strategy is to calibrate the difficulty of trials to the average ability of the population of interest. For example, consider a task trial with two response options. Given the Bernoulli distribution, the variance in responses on this trial is maximized when the probability of choosing either response is equal (i.e. 50\%/50\%). Thus, aggregating across many trials, between-participants variance is maximized --- and task reliability is improved --- when the difficulty of all items are around the average ability of the sample \cite{gulliksen1945relation, ferrando2007external} (or somewhere above this point when guessing the correct response is an issue \cite{lord1952relation, feldt1993relationship}). Of course, this design principle is only helpful to the degree that an experimenter  precisely knows the average ability level of their participants. If this is unknown or poorly characterized, then it is instead desirable to design a task with trials spanning a uniform range of difficulty.

@@

An alternative approach to improving the reliability of difference scores is simply to avoid using them in the first place. That is, experimenters may opt to design tasks or analyses in such a way to avoid the necessity of difference scores in the first place. This recommendation has a long history in experimental psychology. Indeed, because difference scores will virtually always be less reliable, many authors have advocated to simply abandon them \cite{lord1956measurement, cronbach1970we, edwards2001ten}.

>>>>

Cognitive tasks have already demonstrated, and continue to hold, tremendous promise for biological psychiatry. Variability in performance on cognitive tasks provides experimenters with crucial insights into the relationship between individual differences in cognitive processing and behavioral phenomena outside of the laboratory or clinic (e.g. reward sensitivity in a monetary-incentivized task and anhedonia in depression). Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{spiegel2021relations} and cognitive decline \cite{hartshorne2015does}. Cognitive tasks, then, are invaluable tools for building and refining our understanding of psychiatric symptoms and syndromes. For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

% P2: Definitions + Three key properties of behavioral measures: discriminability, reliability, and validity

We define a cognitive task as any experimental paradigm that measures behavior in order to make inferences about one or more cognitive processes (e.g. Stroop task, delay discounting task, reversal learning task). Cognitive-task measures for characterizing individual differences are descriptive or model-based indicators of behavioral performance such as proportion correct, average response time, and learning rate. The psychometric quality of such a measure can be summarized by three key properties: \emph{discriminating power}, \emph{validity} and \emph{reliability} \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to measure variability in participants' performance. This is a necessary property of tasks to be used for studying individual differences; where there is no variation in performance, there are no individual differences to study. The validity of a task measure concerns whether it actually measures what it purports to measure. Finally, the reliability of a task measure characterizes the degree to which it consistently rank-orders participants across measurement instances. Put another way, a task measure is reliable if, assuming participants have not changed, it produces the same score for each participant over time. The focus of this review is task reliability, which is a prerequisite for validity.

% P3: definition and Importance of reliability
\subsubsection*{The formal definition of reliability}

In classical test theory, the variance in observed scores on a task measure $X$ is the sum of the true score variance $T$, reflecting real individual differences in the latent construct of interest, and measurement error $E$, i.e., $X = T + E$. The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance: $\rho = \frac{T}{T + E}$. Thus, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences rather than reflecting noise. Reliability is therefore a prerequisite for validity: an unreliable task measure reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g. task performance and self-reported symptom score) is bounded by their individual reliabilities \cite{Spearman1904-mo}: 
\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}
where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As all reliabilities are $<1$, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (Figure \ref{fig:fig01}). As an important corollary, as measure reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

\begin{figure}[tp]
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig01.png}
    \caption{\small The relationship between measure reliability, observed correlations, and statistical power. (A) The maximum expected observed correlation between two measures as a function of their true (latent) correlation and reliability. As the reliability of two measures decreases, so too does their observed correlation. (B) Required sample size for 80\% power to detect true correlations between two measures given their reliability. As the reliability of two measures decreases, the number of participants required to detect an association increases. Even with a large sample size of $N=500$, a measure with reliability $\rho_{tt} = 0.5$ will only allow detection of correlations above $0.3$, which are uncharacteristically high in the field.}
    \label{fig:fig01}
\end{figure}

@@

To further complicate matters, the reliability of a task measure is not absolute -- it reflects interactions between task design, the participants, and the context in which the task is administered. Indeed, task reliability can vary as a function of parameters of the experiment such as the specific stimulus set, number of trials, time limits \cite{paap2016role, cooper2017role}); sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in clinic, online); response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); and scoring or estimation method \cite{Rouder2019-am, haines2020learning}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children. Their task performance may drop to chance level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may increase the measurement error, leading to a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. Ideally researchers would investigate and report the reliability of task measures as part of any analysis or, at the very least, after having made changes to a task or scoring procedure or when administering the task to new sample populations or in new testing contexts.

@@

Although verifying the reliability of cognitive task measures is paramount to individual-differences research, the reliability of task measures is seldom reported \cite{Green2016-xw, Parsons2019-jw}. When they are reported, task measures frequently exhibit lower reliability than what is conventionally considered acceptable for individual-differences research ($\rho_{tt}$ on the order of 0.7 - 0.8) and the reliability regularly achieved by self-report measures. Indeed, many studies have now found that task measures exhibit moderate-to-low reliability \cite{Hedge2018-lf, Frey2017-uz, Enkavi2019-oh, Von_Bastian2020-tm, Nitsch2022-pe, verdejo2021unified}.

One possible explanation for this finding is the so-called ``reliability paradox'' of cognitive tasks \cite{Hedge2018-lf}, which states that the often lackluster reliability of tasks is a result of a mismatch in goals between experimental and individual-differences psychological research. In experimental psychology, the goal is often to demonstrate the existence of a behavioral effect. One means of increasing the power to detect an effect is to minimize between-participants variance. This is the exact opposite of what is desirable for individual differences research, where between-participants variance is essential to achieving reliable task measures. Thus, the tendency in biological psychiatry to adopt the most prominent tasks in experimental psychology---the ones that most reliably demonstrate a behavioral effect---may actually hamstring our efforts to study individual differences.

@@

Regardless, we do not believe that task measures are inherently less reliable than self-report measures, and that pessimism about task-based individual-differences research is warranted. It is possible to (re)design tasks to achieve good reliability, even to the high levels dictated by conventional standards \cite{waltmann2022sufficient, sullivan2022enhancing, kucina2022solution, snijder2022psychometric}. The purpose of the current article is to provide a narrative review of approaches to improve task-measure reliability. The article is divided into two main parts. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of approaches for improving task reliability using concrete examples from the published literature. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing task, to achieve sufficient reliability for use in individual-differences research.

@@

There are two measures of reliability of relevance for task measures: internal consistency and test-retest reliability. Internal consistency is the reliability of a measure in a single administration of a task. Test-retest reliability (also known as temporal stability) is the reliability of a task measure across two or more administrations. Even for cross-sectional experiments, where stability over time is of no scientific interest, test-retest reliability is a useful index. This is because estimates of internal consistency tend to be inflated due to between-participants variance from construct-irrelevant, state-dependent factors (e.g. current mood, fatigue). Given sufficient time between testing sessions, test-retest reliability should be less biased by state factors, and thus a better estimate of the true cross-sectional task-measure reliability. However, whereas high internal consistency is always desirable for individual difference studies, high test-retest reliability may not be, depending on the construct a researcher is intending to measure. Low one-month test-retest reliability may not be problematic for an index of a transient cognitive process (e.g., mood-dependent attentional biases), but is a problem if individual differences in a cognitive process are hypothesized to be stable across time (e.g., extraversion and social reward processing).

The test-retest reliability of a task measure can be calculated in numerous ways \cite{Parsons2019-jw}. Perhaps the simplest approach is to compute the Pearson correlation between participants' scores from two sessions. An alternative approach is to calculate the intraclass correlation coefficient (ICC), which decomposes a task measure into true score variance and error variance. There are many formulas for calculating ICC \cite{Koo2016-fz}, with the critical distinction being whether reliability is based on the \textit{consistency} or \textit{absolutely agreement} of a task measure across two administrations. Consistency-based ICCs are affected only by the relative ordering of participants across time; that is, they are insensitive to systematic changes to the actual values of a task measure across time (e.g., due to practice effects on task performance). In contrast, absolute-agreement based ICCs measure the degree to which scores are stable across time. The type of ICC to use depends on the experimenter's goals and the ultimate use of the task measure. % desiderata from the measured individual differences. %If an experimenter cares only about the rank-ordering of participants across time, a consistency-based ICC (or Pearson correlation) is preferable. Conversely, if systematic differences in a task measure across task administrations is a concern, then an absolute agreement-based ICC should be preferred.

Calculating the internal consistency of a task measure is more complicated. The most common measure of internal consistency, Cronbach's $\alpha$, requires certain assumptions that are unrealistic for many task data (e.g., equivalence of trials, uncorrelated measurement error). As such, internal consistency for task measures is instead usually calculated via split-$K$ reliability, where $K$ indicates the number of splits. Split-half reliability ($K=2$) is most common. A critical challenge in calculating split-half reliability is in deciding how to partition the data, as estimates of reliability may be biased when partitions are not equally good measures of a behavior \cite{Green2016-xw, pronk2022can}. For example, first-second splitting (i.e., partitioning the data into the first and second halves of an experiment) may underestimate reliability due to differences between the partitions due to practice, fatigue, or other linear time effects. In contrast, odd-even splitting (i.e., partitioning the data into odd and even trials) may be confounded when behavior across trials is non-independent (i.e., measurement error is correlated across trials). Under such conditions, odd-even splitting will artificially inflate the similarity of data across partitions, thereby decreasing estimates of measurement noise and overestimating reliability. Therefore, where possible, a permutation-based approach to calculating split-half reliability is recommended \cite{pronk2022can, Parsons2019-jw}. Here, reliability is averaged across many thousands of random partitions of the data into halves. For task measures derived from cognitive models, however, it may be prohibitively computationally intensive to employ such an approach due to the need to re-estimate the task measure for each new subset of the data. Moreover, for learning tasks that are commonly used in computational psychiatry, cognitive model-based task measures cannot be estimated from only a subset of the trials due to inherent non-independence of task behavior across trials. In these cases, one can design tasks with at least two independent blocks. Then a model is fit to each block independently and reliability is calculated using the model parameters estimated from each.

As a final point, traditional sum or mean score estimates of performance (e.g., proportion correct responses, mean response time) may substantially underestimate task reliability in cases where some response variability across trials can be accounted for rather than treated as noise \cite{Rouder2019-am, haines2020learning}. These summary scores are noisy (and thus less reliable) estimates of task performance as they neglect trial-by-trial variability. Instead, it is often preferable to use generative or statistical models that account for trial-level variability in behavior. By directly modeling trial-by-trial variability, estimates of participants' performance are effectively de-noised. It is also advisable to use hierarchical models that partially pool data across participants, which in turn further increases the precision of estimates of participants' performance. Together, these methods can substantially boost estimates of reliability \cite{Rouder2019-am, haines2020learning}, as multiple studies have demonstrated  \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}. For a detailed discussion of generative models in the context of task reliability, see Haines (this issue).

@@

As defined above, the reliability of a task measure is the proportion of variance attributable to between-participant differences relative to measurement error. Thus, the two major strategies for improving the reliability of a measure are to increase between-participant variability or decrease measurement error. In what follows, we discuss approaches for accomplishing each objective in turn. Where appropriate, we highlight studies that are exemplary for improving the reliability of a task measure by implementing a particular strategy.

@@

By definition, the reliability of a task measure is zero when there is no variability across participants. Thus, range restriction of task measures via ceiling or floor effects is a serious obstacle to reliability. Siegelman and colleagues \cite{siegelman2017measuring} noted the consequences of floor effects on reliability in the context of statistical learning tasks. In such tasks, participants must learn to identify subtle patterns in the transition probabilities underlying a continuous sequence of stimuli. In reanalyzing archival datasets, Siegelman and colleagues found a majority of participants were at chance-level performance in discriminating between legitimate and foil sequence patterns; consequently, the reliability of conventional proportion correct measures suffered. In response, the authors designed a new statistical learning task involving stimulus sequences that ranged more widely in their difficulty to learn. Only a minority of participants showed chance-level performance on this new task and, as such, the reliability of proportion correct scores improved. Similarly, in developing an abbreviated working-memory task, Oswald and colleagues \cite{oswald2015development} found that they could remove the easiest trials---those with ceiling level performance---with virtually no change to task reliability. This is because those trials are incapable of differentiating ability across participants and therefore cannot contribute significantly to the reliability of the task.

Researchers administering a task to a new population should be especially wary of range-restriction effects. Cognitive tasks calibrated for one group of participants may not be adequately sensitive for others due to being too easy or difficult for a different group. For example, Arnon and colleagues \cite{arnon2020current} found that statistical learning tasks developed for adults were too difficult for young children and therefore yielded unreliable discrimination scores in that population. Similarly, Kyllonen and colleagues \cite{kyllonen2019general} were motivated to develop a battery of fluid-reasoning measures for highly educated adults due to observed ceiling effects in performance when using preexisting fluid-reasoning tasks in this population.

@@

A related issue for task reliability is practice effects, where participants' performance on a task improves with repeated administrations. Practice effects are relatively common for cognitive tasks \cite{hausknecht2007retesting, scharfen2018retest}. They might occur due to the attenuation of task-irrelevant nuisance factors (e.g. performance anxiety) and/or the learning of task-specific knowledge or strategies. Practice effects are not inherently an issue for reliability --- especially if an experimenter is only interested in the consistency, but not the absolute agreement, of participants' performance over time --- but they can become a pernicious issue if they are severe enough to induce ceiling effects. For example, Paredes and colleagues \cite{paredes2021psychometric} observed large practice effects on the Pavlovian go/no-go task for short retest intervals (3 days, 14 days), which resulted in poor estimates of test-retest reliability. In developmental and lifespan studies, practice effects are potentially complicated by their interaction with age \cite{anokhin2022age, salthouse2010influence}; that is, practice-induced ceiling effects may present in some age groups but not others.

One strategy for minimizing practice effects is simply to increase the time interval between task administrations. The more time that elapses between sessions, the greater the probability that participants will have forgotten task-specific knowledge or strategies \cite{hausknecht2007retesting, scharfen2018retest}. Of course, this solution may not always be possible or desirable, especially if a researcher is only able to or specifically interested in studying a behavior over a short time period. Moreover, some forms of learning do not easily dissipate with time.

A second strategy is to design tasks so as to prevent or discourage the formation of task-specific strategies. For example, McLean and colleagues \cite{mclean2018towards} investigated the repeatability of the beads task. In the beads task, participants are presented with two jars containing beads of two colors in equal but opposite ratios. In each trial of the task, a predetermined sequence of beads is drawn from one jar. Participants must decide which jar beads were being drawn from or request to see more beads. In a typical version of the task, the same sequence of bead draws is used across all trials. McLean and colleagues found that participants were aware that the sequence repeated across trials and, as a consequence, became more erratic in their decision to witness more bead draws with additional trials. In response, the authors developed a new version of the task that included distractor sequences of bead draws. This new design was effective in preventing participants from becoming aware of the target sequence, which in turn resulted in more consistent responding, which improved the reliability of participants' information seeking scores.

@@

The preceding sections described potential threats to between-participants variability, but not approaches to improve it. A primary strategy for increasing between-participants variability is to enhance the experimental effect. Amplifying the magnitude of an experimental effect (e.g., making a task more challenging, increasing the potency of affect induction) typically increases the range of participants' responses to it. For example, Kucina and colleagues \cite{kucina2022solution} investigated the reliability of cognitive conflict effects in new versions of several standard cognitive-control tasks (e.g., Stroop, Flanker, Simon) that amplified cognitive interference via two task design features. First, they combined multiple sources of cognitive interference in the same task, for example by combining the Stroop and Simon effects to create a ``Stroopon'' task. Second, for a subset of trials, they required participants to make multiple responses based on both relevant and irrelevant stimuli attributes. Compared to the standard version of these tasks, these experimental manipulations had the effect of increasing task demands, which resulted in greater between-participants variance, and higher reliability as a consequence. (See also Snijder and colleagues \cite{snijder2022psychometric} for a similar redesign of classic cognitive-control tasks that improved reliability by increasing proactive control demands.)

% Similarly, Snijder and colleagues \cite{snijder2022psychometric} studied the reliability of cognitive conflict effects in multiple redesigned cognitive-control tasks (e.g., Stroop, AX-CPT, Cued Task Switching, Sternberg) that required of participants either more proactive control (i.e., maintained vigilance) or more reactive control than the traditional versions of these tasks. The authors found that the proactive control versions of the task were more demanding, which similarly led to larger cognitive interference effects and better task reliability.

A related strategy is to calibrate the difficulty of the task to the average ability of the population of interest. For example, consider a task trial with two response options. Given the Bernoulli distribution, the variance in responses on this trial will be maximal when the probability of choosing either response is equal. Thus, aggregating across many trials, between-participants variance is maximized --- and task reliability is improved --- when the difficulty of all items is matched to the average ability of the sample \cite{gulliksen1945relation, ferrando2007external} (or slightly higher if participants can guess the correct response \cite{lord1952relation, feldt1993relationship}). Of course, this design principle is only helpful to the degree that a researcher knows the average ability level of their participants. If this is unknown or poorly characterized, then it is instead desirable to design a task with trials spanning a uniform range of difficulty.

@@

Rather than improving the reliability of difference scores, one can simply avoid using them in the first place. This recommendation has a long history in experimental psychology. Indeed, because difference scores will virtually always be less reliable, many authors have advocated to abandon them \cite{lord1956measurement, cronbach1970we, edwards2001ten}.
